\chapter{Categorical predictor variables}



\section{Dummy coding}
As we have seen in Chapter 1, there are largely two different types of variables: quantitative variables and qualitative variables. Quantitative variables say something about how much of an attribute is in an object: for instance height (measured by inches) or heat (measured in degrees Kelvin). Qualitative variables say something about the quality of an attribute: for instance colour (red, green, yellow) or placement (aisle seat, window seat).

In the chapters on simple and multiple regression we have seen that both the dependent and the independent variables were all quantitative. The linear model used in regression analysis always involves a quantitiative dependent variable. However, in such analyses it is possible to use qualitative independent variables. In this chapter we explain how to do that and how to interpret the results. 

The basic trick that we need is \textit{dummy coding}. Dummy coding involves making one or more new variables, that reflect the different categories of a qualitative variable. First we focus on qualitative variables with only two variables. Later in this chapter, we will explain what to do with qualitative variables with more than two categories. 

Suppose we have two different placements in a bus: aisle seats and window seats. Suppose we ask 5 people who have travelled from Amsterdam to Paris during the last 12 months, whether they had an aisle seat or a window seat, and how much they payed for the trip. Suppose we have the variables, person, seat and price. Table \ref{tab:dummy_1} shows part of the anonymized data.

\begin{kframe}


{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in rnorm(5, 60, 5) \%>\% round(0): could not find function "{}\%>\%"{}}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in data.frame(person, seat, price): object 'price' not found}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in data.bus \%>\% head() \%>\% xtable(caption = "{}Bus trips to Paris."{}, : could not find function "{}\%>\%"{}}}\end{kframe}


With dummy coding, we make a new variable that only has values 0 and 1, and that conveys the same information as the \textbf{seat} variable. The resulting variable is called a dummy variable. Let's call this dummy variable \textbf{window} and give it the value 1 for all persons that travelled in a window seat. We give the value 0 for all persons that travelled in an aisle seat. We can also call the new variable \textbf{window} a boolean variable with TRUE and FALSE, since in computer science, TRUE is coded by a 1 and FALSE by a 0. Whatever you want to call it, the data matrix including the new variable is displayed in Table \ref{tab:dummy_2}.

\begin{kframe}


{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in data.frame(person, seat, window, price): object 'price' not found}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in data.bus \%>\% head() \%>\% xtable(caption = "{}Bus trips to Paris."{}, : could not find function "{}\%>\%"{}}}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(predvars, data, env): object 'price' not found}}\end{kframe}





